### 2016-10-25 doc/publications/plos16/plos16.tex
General fixes

\documentclass[10pt,a4paper]{article}  %\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

@@

\usepackage{footmisc}

@@

\newcommand{\cursedforest}{{\sc CursedForest}}
\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

@@

\usepackage{multirow}

@@

% Title must be 250 characters or less.
% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
\begin{flushleft}
{\Large
\textbf\newline{Cursed Forest - A random forest implementation for ``big'' and ``wide'' data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Aidan O'Brien\textsuperscript{1},
Piotr Szul\textsuperscript{2},
Stephanie Li\textsuperscript{3},
James Doecke\textsuperscript{3},
Nick Ellis\textsuperscript{4},
Robert Dunne\textsuperscript{5}, and
Denis C. Bauer\textsuperscript{1,*}
%, with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\bf{1} Health \& Biosecurity, CSIRO, Sydney, NSW, Australia
\\
\bf{2} Data61, CSIRO, Brisbane, QLD, Australia
\\
\bf{3} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
\\
\bf{4} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
\\
\bf{5} Data61, CSIRO, Sydney, NSW, Australia
\\
\bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
\tableofcontents
\clearpage
\section{Notes}
\begin{itemize}
\item   I think the structure of the first section should be
  \begin{itemize}
  \item  convergence results from Biau 2012  (done)
  \item consider  \cite{Segal.2004} examples
  \item simulations reproduced from Genuer et al 2010 (no -- we need classification examples)
  \item check Piotr's simulation  (done)
  \item run sparkML or Cursed forest on the  simulation data 
  \item we extend the simulations to many more variables 
  \item I dont think I can get R to handle chromosme 1 etc. May have to run CF on chromosme 22, see table
    \ref{table:smaller_data} -- On it, Aidan
  \item variable  selection may succeed where accurate prediction does not
  \item discussion of the effect of the ntree and mtry parameters for variable selection with many noise variables
  \item   perhaps we argue that in the case where the relationship of $X$ and $y$ takes a particular form (step, ramp) we can recover
    the significant variables in the presence of a very large number of noise variables
  \item in the case of genomic data, our functions of interest are more likely to be stop or ramp functions rather than harmonic
    functions. So we have some hope.
  \end{itemize}
\item check out 
  \begin{itemize}
  \item D\'iaz-Uriarte, R., \& Alvarez de Andr\'es, S. (2006). (done)
  \item \cite{Goldstein.et.al.2011}
notes that the 0, 1, 2 encoding of alleles is an additive model i.e. each additional minor allele increases the effect 
\begin{verbatim}
Different genetic effects
Type	Mechanism	Partition
Additive	Each additional minor allele increases variation	0, 1, 2
Dominant	Presence of at least 1 minor allele increases variation	0, 1/2
Recessive	Two minor alleles needed for variation	0/1, 2
Heterosis	Heterozygote leads to variation	0/2, 1
\end{verbatim}

@@

\section{Abstract}
The abstract.

@@

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forest specifically designed to cater for
``big'' (many samples) and ``wide'' (many features) data sets. \cursedforest\ is capable of parallelising the split for
each node in a tree thereby handling millions of features, such as in the 1000 Genome data, which consists
of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also utilising Spark to read in and
manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms existing tools on even small
datasets where multi-threading generally performs well. 
\marginpar{``performant'' -- a person who performs, an actor or celebrant }
 \marginpar{Piotr this is your 20 trees has the important
  feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be
  a result section.}
Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees,
which boosts the accuracy for variable importance selection~\cite{Szymczak2016}.

@@

%% We therefore developed \cursedforest\, a Spark-based random forest algorithm able to process big and ``wide'' data.
%% % Unlike the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
%% % excessive resource requirements on the driver or worker nodes.  DENIS Piotr this is your 20 trees has the important feature in
%% % the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.
%% \marginpar{Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top
%%   element, do we need to spell this out better, it should be a result section.}  Harnessing the virtually unlimited
%% capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for
%% variable selection as demonstrated by {\sc r2VIM}~\cite{Szymczak2016} on smaller datasets.

@@

%Spark ML's RF implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest dimensionality.  
%Although Spark ML can build a RF model on
%a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
%and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).  This
%unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
%scaling to larger datasets. The memory requirements per executor also increases with dimensionality due to the data types
%Spark ML uses, which we elaborate on in the discussion.
%
%We therefore developed \cursedforest\, a Spark-based random forest algorithm able to process big and ``wide'' data. 
%Unlike
%the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without
%excessive resource requirements on the driver or worker nodes.
%DENIS Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.

\marginpar{Needs to be rewritten once we actually have the sections} 
In the first section we explore some of the
properties of random forests, demonstrating that any potential flaws in CursedForest are an inherent property of Random
Forests. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability of CursedForest in
respect to the dimensionality of the data, building an RF model on whole-genome data from the 1000 Genomes Project. Here
we report on the accuracy (in terms of the ARI)
\marginpar{what is ARI?}
 and the error.  Finally, given the role different parameter values can
play in model construction, we explore the effect tuning these parameters can have on the prediction accuracy of the
model.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Cursed Forest.}
% DENIS: Aidan and Piotr can you please explain here how Cursed forest is implemented
Traditionally, VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data
in a ``DataFrame'', similar to DataFrames in R.

@@

On the other hand, CursedForest (CF) is specifically designed to handle wide ``cursed'' data. It avoids the relation
between memory and dimensionality by avoiding calculations that rely on entire feature vectors.

Compared to Spark ML, CursedForest further parallelizes work down to individual features.  For example, for each node of
a tree, CursedForest will distribute tasks that consist of single features (variants), for every individual.  Each of
these tasks will calculate the information gain for that specific feature.  Once these tasks have completed, the results
are reduced to return the feature which gives the greatest information gain.  This process is then repeated until
CursedForest has calculated the entire decicion tree.

The current implementation of Cursed Forest uses:
\begin{itemize}
\item an ``information gained'' criteria for splitting. Let $f_q$ be the fraction of items labeled with value $q$ where $q= 1,
  \ldots, Q$ at a node. The  entropy of the node is $H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i.$
The expected information gain is the change in entropy $H$ from a prior state $T$ to a state that takes some
information $a$ as given: 
\[ IG(T,a) = H(T) - H(T|a). \]
In this case, the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children)
\item variable importance is the sum of IG over all notes and all trees for that variable
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
%DENIS: Aidan can you please fill this in of where we got it from and how we process it for all the different tools to use
\subsubsection{1000 Genomes}
%DENIS: Rob how was the synthetic data generated
\subsubsection{Synthetic data}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{CursedForest can evaluate variable importance for millions of features  }\label{synthetic}
%In this section we demonstrate that \cursedforest 's ability to rapidly grow large number of trees, \texttt{ntree}, in
%parallel results in a stable feature importance ranking.  This concept was originally proposed by Liaw and
%Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to
%run, the ranking of the variables will stabilise with growing \texttt{ntree}.
In this section we demonstrate CursedForrest two main features, namely variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.
%We therefore generate a synthetic dataset with varying numbers of variables (features) and samples, as discussed in the method section. 
We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000 samples, as discussed in the
method section.  Briefly we generate features by randomly sampling from a three value vector (1,2,3) with equal
probabilities, which mimics genomic mutation data and corresponds to homozygote reference, heterozygote or homozygote
alternative.  We then generate the response variable as a linear combination of 5 weighted features, modulating a
complex trait of multiple interacting features.

@@

From the synthetic data we can conclude that CursedForest is able to perform accurate feature selection as well as
accurate classification for large numbers of samples with large feature vectors.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{CursedForest selects previously demonstrated cancer variants }
%Feature selection can be useful for determining variables the contribute to the label or for perhaps selecting the n most important
%variables for further analysis. 
As established in the pervious section, CursedForest allows the user to retrieve a subsection of features, ranked by
their importance to identify the variables predictive of the label or selecting the $n$ most important variables for
further analysis.  We demonstrate this feature on The Cancer Genome Atlas (TCGA) BRCA dataset.  We build this model on
VCF files retrieved from the TCGA data portal.  For the label, we use the cell type (i.e. tumor or normal).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 3: Classification selection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{CursedForest outperforms existing methods for classification problems}
As established in Section~\ref{synthetic}, CursedForest 
In this section we compare \cursedforest 's performance against other published methods for predicting ethnicity on the
1000 genomes dataset.

\begin{table}[!ht]
\begin{minipage}{\textwidth}
\begin{adjustwidth}{-1.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\caption{
{\bf Performance comparison between the different machine learning algorithms.}}
\begin{tabular}{|l+l|l|l|l|l|l|p{1cm}|}
\hline
%\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}                               \\ \hline
\bf{Genome \%}                      & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/Node} \\
\hline

\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest  & 0.077 (XX) &                  &                                                                   \\
                                                & RF - Spark ML &            &                  &                                                                   \\
                                                & RF - R        & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\
                                                & ranger -R     & 0.94       & 1 min 10 seconds & 16GB  (one processor)                                             \\
                                                & ranger c++    & 0.94       & 7 mins           & --nthreads 20                                                     \\

@@

                                                    & ranger  -R    & 0.94       & 4 1/2 mins       & 21GB (one processor)                                                                \\

@@

\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & CursedForest  & 0.063 (XX) & 1hr 58min        & 4GB                                                                                 \\

@@

                                                    & ranger  R     & 0.95       & 28 mins          & num.threads=10\footnote{don't

@@

\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & CursedForest  & 0.046 (XX) & 4hr 0min                                             & 4GB         \\

@@

                                                    & ranger  R     &  0.97      & 58 mins\footnote{ but took 6 hours to load the

@@

& ranger   -R        &             0.973     &      03:58           &       pearcey --cpus-per-task=15 --mem 1600gb   906GB \\

@@

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
& RF - Spark ML & - & - & - \\
& LR - Spark ML & - & - & - \\ 
& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ 
& ranger (C++)       &        NA     &        NA     &             80 CPU -- runs out of memory (12.8GB per CPU) \\
& ranger (C++)       &        NA     &        NA      &            160
                                                        CPU -- runs
                                                        out of time at
                                                        20 hours \\

% \begin{tabular}{|p{1.2cm}|p{1cm}|p{1.5cm}+l|l|p{1.8cm}|p{1.5cm}|}
% \hline
% %\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
% \bf{Data- set} & \bf{Samp- les} & \bf{Feat- ures}  & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/ Node} \\
% \hline
% %\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest & 0.077 (XX)  &  &  \\
% phase1 chr22 &1,092 & 490,036 & CursedForest & 0.077 (XX)  & 9min 16sec & 1GB\footnote{\label{note10}10 nodes} \\
% &&& RF - Spark ML & 0.069 (0.89) & 14min 13sec & 1GB\footref{note10}  \\
% &&& RF - R &  0.05  & 24min  & 18GB\footref{note1}\footnote{using the \texttt{parallel} packages}\\
% &&& ranger -R &         0.05 &        1min 10sec  &          16GB  (one processor) \\
%  \hline
% phase3 chr22 & 2,504 & 1,103,548 &  &  &  &  \\
% &&& CursedForest & 0.053 & 9 min 8 sec & 2GB\footnote{\label{note20}20 nodes}\\
% &&& RF - Spark ML & 0.058 (0.92) & 1hr 1min & 2GB\footref{note20} \\
% &&& RF - R & NA  & NA & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface}\\
% &&&  H2O   &        0.05 &     16 hours       &   600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
% \hline
% phase3 chr1 & 2,504 & 6,450,364 & CursedForest & 0.063 (XX) & 1hr 58min & 4GB \\
% &&& RF - Spark ML & (0.94) & 8hr 29min & 8GB \\
% &&& LR - Spark ML&  & 2hr 6min & 8GB\\ 
% &&& ranger  R   &  0.06   &     28 mins  &            num.threads = 10\footnote{don't know how to get the memory used, This
%   was done on dpas-03} \\ 
% \hline
% phase3 chr1-3 & 2,504 & 19,328,051  & CursedForest & 0.046 (XX) & 4hr 0min & 4GB\\
% &&& RF - Spark ML & - & -  & -\\
% &&& LR - Spark ML & & 14hr 48min & 8GB \\ 
% &&& ranger  R   &  0.05   &     58 mins\footnote{ but took 6 hours to load the data}  &      909GB\footnote{I think this is total memory. The job was run with
%   ntasks-per-node=160 cores-per-socket=10 on ruby }  \\ 
% \hline
% phase3 chr1-22 & 2,504 & 81,047,467 & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
% &&& RF - Spark ML & - & - & - \\
% &&& LR - Spark ML & - & - & - \\ 
% %&&& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ 
% \hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project.
Each random forest model consists of 50 trees.
\end{flushleft}
\label{table1}
\end{adjustwidth}
\end{minipage}
\end{table}

@@

In this section we explore the performance of CursedForest by testing its ability to scale to different sizes of data and computational resources.

In order to asses these characteristics we ran CursedForst classification on synthetics datasets with varying numbers of variables (features) and samples, similar to the dataset used in to evaluate ranger [REF: ranger], allocating varying number of CPU cores to the CursedForest and also varying computational complexity of random forest by using range of mtry values.

@@

To conclude CursedForst is capable of processing datasets with 50M variables and 10K samples, which is the order of magnitude of the entire genome variants expected in such dataset, for which it can build approximately 60 trees per hour.

>>>>

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

@@

@@

\newcommand{\cursedforest}{{\sc CursedForest}}

@@

%% Additional packages
\usepackage{footmisc}
\usepackage{multirow}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{\cursedforest - A random forest implementation for ``big'' and ``wide'' data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Aidan O'Brien\textsuperscript{1},
Piotr Szul\textsuperscript{2},
Stephanie Li\textsuperscript{3},
James Doecke\textsuperscript{3},
Nick Ellis\textsuperscript{4},
Robert Dunne\textsuperscript{5}, and
Denis C. Bauer\textsuperscript{1,*}
%, with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Health \& Biosecurity, CSIRO, Sydney, NSW, Australia
\\
\textbf{2} Data61, CSIRO, Brisbane, QLD, Australia
\\
\textbf{3} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
\\
\textbf{4} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
\\
\textbf{5} Data61, CSIRO, Sydney, NSW, Australia
\\
\bigskip

@@

\end{flushleft}
\tableofcontents
\clearpage
\section{Notes}
\begin{itemize}
\item   I think the structure of the first section should be
  \begin{itemize}
  \item  convergence results from Biau 2012  (done)
  \item consider  \cite{Segal.2004} examples
  \item simulations reproduced from Genuer et al 2010 (no -- we need classification examples)
  \item check Piotr's simulation  (done)
  \item run sparkML or CursedForest on the  simulation data 
  \item we extend the simulations to many more variables 
  \item I dont think I can get R to handle chromosme 1 etc. May have to run CF on chromosme 22, see table
    \ref{table:smaller_data} -- On it, Aidan
  \item variable  selection may succeed where accurate prediction does not
  \item discussion of the effect of the ntree and mtry parameters for variable selection with many noise variables
  \item   perhaps we argue that in the case where the relationship of $X$ and $y$ takes a particular form (step, ramp) we can recover
    the significant variables in the presence of a very large number of noise variables
  \item in the case of genomic data, our functions of interest are more likely to be stop or ramp functions rather than harmonic
    functions. So we have some hope.
  \end{itemize}
\item check out 
  \begin{itemize}
  \item D\'iaz-Uriarte, R., \& Alvarez de Andr\'es, S. (2006). (done)
  \item \cite{Goldstein.et.al.2011}
notes that the 0, 1, 2 encoding of alleles is an additive model i.e. each additional minor allele increases the effect 
\begin{verbatim}
Different genetic effects
Type	Mechanism	Partition
Additive	Each additional minor allele increases variation	0, 1, 2
Dominant	Presence of at least 1 minor allele increases variation	0, 1/2
Recessive	Two minor alleles needed for variation	0/1, 2
Heterosis	Heterozygote leads to variation	0/2, 1
\end{verbatim}

@@

% Please keep the abstract below 300 words
\section*{Abstract}
The abstract.

@@

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forest specifically designed to cater for
``big'' (many samples) and ``wide'' (many features) data sets. \cursedforest\ is capable of parallelising the split for
each node in a tree thereby handling millions of features, such as in the 1000 Genomes Project data, which consists
of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also utilising Spark to read in and
manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms existing tools on even small
datasets where multi-threading generally performs well. 
\marginpar{``performant'' -- a person who performs, an actor or celebrant }
 \marginpar{Piotr this is your 20 trees has the important
  feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be
  a result section.}
Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees,
which boosts the accuracy for variable importance selection~\cite{Szymczak2016}.

@@

%% We therefore developed \cursedforest\, a Spark-based random forest algorithm able to process big and ``wide'' data.
%% % Unlike the standard Apache implementation, \cursedforest\ is able to easily process data with millions of dimensions without
%% % excessive resource requirements on the driver or worker nodes.  DENIS Piotr this is your 20 trees has the important feature in
%% % the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.
%% \marginpar{Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top
%%   element, do we need to spell this out better, it should be a result section.}  Harnessing the virtually unlimited
%% capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for
%% variable selection as demonstrated by {\sc r2VIM}~\cite{Szymczak2016} on smaller datasets.

@@

%Spark ML's RF implementation is not able to handle the extremely ``wide'' genomic data as it was developed for large number of samples with only modest dimensionality.  
%Although Spark ML can build a RF model on
%a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated
%and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}).  This
%unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless
%scaling to larger datasets. The memory requirements per executor also increases with dimensionality due to the data types
%Spark ML uses, which we elaborate on in the discussion.
%
%We therefore developed \cursedforest\, a Spark-based random forest algorithm able to process big and ``wide'' data. 
%Unlike
%the standard Apache implementation, \cursedforest\ is able to easily process data with millions of dimensions without
%excessive resource requirements on the driver or worker nodes.
%DENIS Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.

\marginpar{Needs to be rewritten once we actually have the sections} 
In the first section we explore some of the
properties of random forests, demonstrating that any potential flaws in \cursedforest\ are an inherent property of Random
Forests. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability of \cursedforest\ in
respect to the dimensionality of the data, building an RF model on whole-genome data from the 1000 Genomes Project. Here
we report on the accuracy (in terms of the ARI)
\marginpar{what is ARI?}
 and the error.  Finally, given the role different parameter values can
play in model construction, we explore the effect tuning these parameters can have on the prediction accuracy of the
model.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest.}
% DENIS: Aidan and Piotr can you please explain here how CursedForest is implemented
Traditionally, VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data
in a ``DataFrame'', similar to DataFrames in R.

@@

On the other hand, \cursedforest\ (CF) is specifically designed to handle wide ``cursed'' data. It avoids the relation
between memory and dimensionality by avoiding calculations that rely on entire feature vectors.

Compared to Spark ML, \cursedforest\ further parallelizes work down to individual features.  For example, for each node of
a tree, \cursedforest\ will distribute tasks that consist of single features (variants), for every individual.  Each of
these tasks will calculate the information gain for that specific feature.  Once these tasks have completed, the results
are reduced to return the feature which gives the greatest information gain.  This process is then repeated until
\cursedforest\ has calculated the entire decicion tree.

The current implementation of \cursedforest\ uses:
\begin{itemize}
\item an ``information gained'' criteria for splitting. Let $f_q$ be the fraction of items labeled with value $q$ where $q= 1,
  \ldots, Q$ at a node. The  entropy of the node is $H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i.$
The expected information gain is the change in entropy $H$ from a prior state $T$ to a state that takes some
information $a$ as given: 
\[ IG(T,a) = H(T) - H(T|a). \]
In this case, the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children)
\item variable importance is the sum of IG over all notes and all trees for that variable
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
%DENIS: Aidan can you please fill this in of where we got it from and how we process it for all the different tools to use
\subsubsection{1000 Genomes Project}

We obtained the 1000 Genomes Project data as VCF files from their FTP site.

%DENIS: Rob how was the synthetic data generated
\subsubsection{Synthetic data}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
%In this section we demonstrate that \cursedforest's ability to rapidly grow large number of trees, \texttt{ntree}, in
%parallel results in a stable feature importance ranking.  This concept was originally proposed by Liaw and
%Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to
%run, the ranking of the variables will stabilise with growing \texttt{ntree}.
In this section we demonstrate \cursedforest\ two main features, namely variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.
%We therefore generate a synthetic dataset with varying numbers of variables (features) and samples, as discussed in the method section. 
We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000 samples, as discussed in the
method section.  Briefly we generate features by randomly sampling from a three value vector (1,2,3) with equal
probabilities, which mimics genomic mutation data and corresponds to homozygote reference, heterozygote or homozygote
alternative.  We then generate the response variable as a linear combination of 5 weighted features, modulating a
complex trait of multiple interacting features.

@@

From the synthetic data we can conclude that \cursedforest\ is able to perform accurate feature selection as well as
accurate classification for large numbers of samples with large feature vectors.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ selects previously demonstrated cancer variants }
%Feature selection can be useful for determining variables the contribute to the label or for perhaps selecting the n most important
%variables for further analysis. 
As established in the pervious section, \cursedforest\ allows the user to retrieve a subsection of features, ranked by
their importance to identify the variables predictive of the label or selecting the $n$ most important variables for
further analysis.  We demonstrate this feature on The Cancer Genome Atlas (TCGA) BRCA dataset.  We build this model on
VCF files retrieved from the TCGA data portal.  For the label, we use the cell type (i.e. tumor or normal).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 3: Classification selection
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ outperforms existing methods for classification problems}
As established in Section~\ref{synthetic}, \cursedforest\ 
In this section we compare \cursedforest 's performance against other published methods for predicting ethnicity on the
1000 Genomes Project dataset.

\begin{table}[!ht]
\begin{minipage}{\textwidth}
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Performance comparison between the different machine learning algorithms.}}
\begin{tabular}{|l+l|l|l|l|l|l|p{1cm}|}
\hline
\bf{Genome \%}                      & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/Node} \\
\hline

\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.077 (XX) &                  &                                                                   \\
                                                & RF - Spark ML  &            &                  &                                                                   \\
                                                & RF - R         & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\
                                                & ranger - R      & 0.94       & 1 min 10 seconds & 16GB  (one processor)                                             \\
                                                & ranger - c++     & 0.94       & 7 mins           & --nthreads 20                                                     \\

@@

                                                    & ranger - R    & 0.94       & 4 1/2 mins       & 21GB (one processor) \\

@@

\multirow{3}{*}{phase3\_chr1: 2,504 x 6,450,364}    & \cursedforest\ & 0.063 (XX) & 1hr 58min        & 4GB \\

@@

                                                    & ranger - R      & 0.95       & 28 mins          & num.threads=10\footnote{don't

@@

\multirow{3}{*}{phase3\_chr1-3: 2,504 x 19,328,051} & \cursedforest\ & 0.046 (XX) & 4hr 0min                                             & 4GB \\

@@

                                                    & ranger - R      &  0.97      & 58 mins\footnote{ but took 6 hours to load the

@@

& ranger - R        &             0.973     &      03:58           &       pearcey --cpus-per-task=15 --mem 1600gb   906GB \\

@@

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & XX (0.96) & 7hr 14min & 8GB \\
& RF - Spark ML & - & - & - \\
& LR - Spark ML & - & - & - \\ 
& kmeans - Spark ML  & - (0.82)      & 30hr 44min    & 24GB \\ 
& ranger (C++)       &        NA     &        NA     &            80 CPU -- runs out of memory (12.8GB per CPU) \\
& ranger (C++)       &        NA     &        NA     &            160
                                                        CPU -- runs
                                                        out of time at
                                                        20 hours \\
\hline
% \begin{tabular}{|p{1.2cm}|p{1cm}|p{1.5cm}+l|l|p{1.8cm}|p{1.5cm}|}
% \hline
% %\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
% \bf{Data- set} & \bf{Samp- les} & \bf{Feat- ures}  & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/ Node} \\
% \hline
% %\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest & 0.077 (XX)  &  &  \\
% phase1 chr22 &1,092 & 490,036 & CursedForest & 0.077 (XX)  & 9min 16sec & 1GB\footnote{\label{note10}10 nodes} \\
% &&& RF - Spark ML & 0.069 (0.89) & 14min 13sec & 1GB\footref{note10}  \\
% &&& RF - R &  0.05  & 24min  & 18GB\footref{note1}\footnote{using the \texttt{parallel} packages}\\
% &&& ranger -R &         0.05 &        1min 10sec  &          16GB  (one processor) \\
%  \hline
% phase3 chr22 & 2,504 & 1,103,548 &  &  &  &  \\
% &&& CursedForest & 0.053 & 9 min 8 sec & 2GB\footnote{\label{note20}20 nodes}\\
% &&& RF - Spark ML & 0.058 (0.92) & 1hr 1min & 2GB\footref{note20} \\
% &&& RF - R & NA  & NA & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface}\\
% &&&  H2O   &        0.05 &     16 hours       &   600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
% \hline
% phase3 chr1 & 2,504 & 6,450,364 & CursedForest & 0.063 (XX) & 1hr 58min & 4GB \\
% &&& RF - Spark ML & (0.94) & 8hr 29min & 8GB \\
% &&& LR - Spark ML&  & 2hr 6min & 8GB\\ 
% &&& ranger  R   &  0.06   &     28 mins  &            num.threads = 10\footnote{don't know how to get the memory used, This
%   was done on dpas-03} \\ 
% \hline
% phase3 chr1-3 & 2,504 & 19,328,051  & CursedForest & 0.046 (XX) & 4hr 0min & 4GB\\
% &&& RF - Spark ML & - & -  & -\\
% &&& LR - Spark ML & & 14hr 48min & 8GB \\ 
% &&& ranger  R   &  0.05   &     58 mins\footnote{ but took 6 hours to load the data}  &      909GB\footnote{I think this is total memory. The job was run with
%   ntasks-per-node=160 cores-per-socket=10 on ruby }  \\ 
% \hline
% phase3 chr1-22 & 2,504 & 81,047,467 & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
% &&& RF - Spark ML & - & - & - \\
% &&& LR - Spark ML & - & - & - \\ 
% %&&& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ 
% \hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project.
Each random forest model consists of 50 trees.
\end{flushleft}
\label{table1}
\end{adjustwidth}
\end{minipage}
\end{table}

@@

In this section we explore the performance of \cursedforest\ by testing its ability to scale to different sizes of data and computational resources.

In order to asses these characteristics we ran \cursedforest\ classification on synthetics datasets with varying numbers of variables (features) and samples, similar to the dataset used in to evaluate ranger [REF: ranger], allocating varying number of CPU cores to the \cursedforest\ and also varying computational complexity of random forest by using range of mtry values.

@@

To conclude \cursedforest\ is capable of processing datasets with 50M variables and 10K samples, which is the order of magnitude of the entire genome variants expected in such dataset, for which it can build approximately 60 trees per hour.
